Statistically-estimated tree composition for the northeastern United States \\ at the time of European settlement

### 2015-03-18 paper.tex
varisou minor edits to paper text and figures

look through code and labbook to make sure have all details

properly cite Goring stuff and reference details of data collection

@@

%
\begin{abstract}
We present a data product of the estimated composition of tree taxa
at the time of European settlement of the United States and the statistical
methodology used to produce the product. Composition is defined as
the proportion of stems larger than approximately X cm diameter at
breast height {[}Charlie? Simon? is X roughly 10?{]} in 22 taxonomic
groupings, generally at the genus level. The data come from settlement
survey records that provide raw data that are transcribed and then
aggregated spatially, giving count data. The domain is divided into
two regions, western (Indiana through Minnesota) and eastern (Ohio
to Maine). Public Land Survey point data in the western region is
aggregated to a regular 8 km grid, while data in the eastern region,
from Town Proprietor Surveys is aggregated at the township level.
The product is based on a Bayesian statistical model fit to the count
data that estimates composition on a regular 8 km grid. The statistical
model allows us to estimate composition at locations with no data
and to smooth over noise caused by limited counts in locations with
data. Critically, it also allows us to quantify uncertainty in our
composition estimates. We expect this data product to be useful for
understanding the state of vegetation in the northeastern United States
prior to large-scale European settlement. In addition to specific
regional questions, the data product can also serve as a baseline
against which to investigate how forests and ecosystems change after
intensive settlement.
\end{abstract}

@@

Euro-American settlement and subsequent land use change occured in
a time transient fashion across North America, and surveys of vegetation
followed a similar pattern. Early surveys (from 1620 until 1825) in
the northeastern United States were areal surveys at a township level
\citep{cogbill2002forests,thompson2013four}, which we call the Town
Proprietor Survey (TPS). These surveys reporting percent composition
of major tree species using sometimes inconsistent common names. Later
surveys, after the establishment of the Public Land Survey System
(PLSS) - from 1819 to 1904 - were point estimates along a regular
grid, with 1 mile spacing \citep{schulte2001original,bourdo1956review,goring2015composition}.
Survey instructions during the PLSS varied through time, and by point
type. This required the application of spatially varying correction
factors \citep{cogbill2015corrections,goring2015composition} to accurately
assess stem density, basal area and biomass from the early settlement
records.

Logging, agriculture and abandonment has left an indelible mark on
forests in the northeastern United States \citep{foster1998land,goring2015composition,rhemtulla2009legacies,thompson2013four},
however most studies have studied these effects at the state level
or below \citep{rhemtulla2009historical,friedman2005regional}, and
with varying spatial resolution, from townships (36 square miles)
to forest zones of hundreds or thousands of square miles. \citet{goring2015composition}
provide a dataset across the upper Midwest that is resolved to an
8 x 8 km grid size, providing broad spatial coverage at a meso-spatial
scale that can be compared to the Forest Inventory and Analysis products
\citet{gray2012forest}. The limitation of this dataset is that it
provides only estimates of variance within cell for PLSS data in the
western region, and no township estimates of error for regions surveyed
as part of the Town Proprietor Survey.

@@

The raw data are obtained from survey records collated from across
the northeastern U.S. by a number of researchers. For the states of
Minnesota, Wisconsin, Illinois, Indiana, and Michigan (the western
subdomain), data are available at survey point locations and have
been aggregated to a regular 8 km grid in the Albers projection. For
the states of Ohio, Pennsylvania, New Jersey, New York and the six
New England states (the eastern subdomain), data are aggregated at
the township level. There is also data from a single township in Quebec
and a single township in northern Delaware. Data are essentially complete
in Minnesota, Wisconsin and Michigan but data in Illinois and Indiana
represent a sample of the full set of grid cells, with survey record
transcription ongoing. Data for the remaining states are a subset
of the full set of townships. Fig. \ref{fig:domain} {[}Fig. 1 - Lyx
is having a numbering problem{]} shows the domain, indicating the
grid cells and townships with data.

@@

\caption{Spatial domain, with locations with data shown in gray. Locations
are grid cells in western portion and townships in eastern portion.}
\end{figure}

@@

Extensive details on the data are available in \cite{Gori:etal:2015}
and the raw data are available at XXX. The aggregation into taxonomic
groups is primarily at the genus level, but in some cases of monospecific
genera, is at the species level. We model the following 22 taxa plus
an ``other hardwood'' category: Atlantic white cedar (Chamaecyparis
thyoides), Ash (Fraxinus spp.), Basswood (Tilia americana), Beech
(Fagus grandifolia), Birch (Betula spp.), Black gum/sweet gum (Nyssa
sylvatica and Liquidambar styraciflua), Cedar/juniper (what is in
here other than Juniperus and is the Juniperus only Juniperus virginiana),
Cherry (Prunus spp.), Chestnut (Castanea dentata), Dogwood (Cornus
spp.), Elm (Ulmus spp. or Ulmus americana???) , Fir (Abies spp.),
Hemlock (Tsuga canadensis), Hickory (Carya spp.), Ironwood (Carpinus
caroliniana and Ostrya virginiana (anything else?)), Maple (Acer spp.),
Oak (Quercus spp.), Pine (Pinus spp.), Poplar/tulip poplar (Populus
spp. and Liriodendron tulipifera), Spruce (Picea spp.), Tamarack (Larix
laricina), Walnut (Juglans nigra). {[}Simon/Charlie to check my Latin
and whether I've left out anything that falls into a given common
name{]} Note that in several cases (e.g., black gum/sweet gum, ironwood,
poplar/tulip poplar, cedar/juniper), because of ambiguity in the common
tree names used by surveyors, a group represents trees from different
families and even orders. For the western subdomain we do not fit
statistical models for Atlantic white cedar and chestnut as these
have 0 and 5 trees present in the dataset, respectively.

@@

There are approximately 520,000 trees from the western subdomain and
420,000 trees from the eastern subdomain. In each subdomain, oak is
the most common taxon and pine the second most common.

Our domain is a rectangle covering all of the states, in the Albers
projection (NAD 1983 Great Lakes and St. Lawrence Albers) {[}Simon,
please confirm this language is right{]}, with the rectangle split
into 8 km cells, arranged in a 296 by 180 grid of cells, with the
centroid of the cell in the southeast corner located at (-71000, 58000).
For the modeling of the western subdomain we use the western-most
146 by 180 grid of cells. For the modeling of the eastern subdomain
we use the eastern-most 180 by 180 grid of cells and then omit 23
cells in the north and 17 cells in the south. {[}check 23 in N vs
17 in N; also check what is in netCDF product - should be for the
full domain with missing values?{]}

@@

We compared the CAR and SPDE models using cross-validation by holding
out data from the fitting process and assessing the fit of the model
on the held-out data. We used two types of test data:
\begin{enumerate}
\item We held out all the data from 95\% of the cells in Minnesota, with
cells selected at random. This was meant to assess the ability of
the model to interpolate from a sparse set of cells/townships and
mimics the limited data in Illinois and Indiana.
\item We held out 5\% of the trees from all of the trees in the dataset
(leaving aside the held-out Minnesota cells). This was meant to assess
the ability of the model to estimate the composition in cells in which
data were available. 
\end{enumerate}
Finally, in a separate sensitivity analysis we instead left out 80\%
of the cells in Minnesota at random. This variation on the first test
above was meant to indicate whether our model comparison conclusions
would be robust as the digitization process for Illinois and Indiana
progresses and provides us with increasingly dense data.

@@

We ran each model for XXX iterations, retaining YYY after burnin and
subsampling to reduce storage needs.

@@

Table \ref{tab:score_cell_fivepercent} shows predictive ability for
Experiment 1: those cells in Minnesota held out of the fitting process,
for both the posterior mean predictions and the full posterior sample.
Table \ref{tab:coverage_cell_fivepercent} shows performance of the
uncertainty estimates. For the posterior distribution over the predictive
score values, the CAR model generally outperforms the SPDE model,
while for the posterior mean predictions, the SPDE model appears to
outperform the CAR model, but we do not have any uncertainty estimates
for this comparison. Coverage and interval lengths are similar between
the two models. From a practical perspective, based on the difference
in mean absolute error, which is on a readily interpretable scale,
the differences between the models are small.

%
\begin{table}

\caption{Predictive ability based on several predictive score criteria for
the CAR and SPDE spatial models when holding out 95\% of entire cells
of data in Minnesota. Smaller values are better.}

\begin{tabular}{|c|c|c|>{\centering}p{3cm}|>{\centering}p{2.5cm}|>{\centering}p{2.5cm}|}
\hline 
 &
\multicolumn{3}{c|}{{\small{Posterior mean of score}}} &
\multicolumn{2}{c|}{{\small{Score of posterior mean predictions}}}\tabularnewline
\hline 
\hline 
 &
{\small{CAR model}} &
{\small{SPDE model}} &
{\small{Posterior Prob. CAR < SPDE}} &
{\small{CAR model}} &
{\small{SPDE model}}\tabularnewline
\hline 
{\small{Brier}} &
{\small{84997}} &
{\small{86803}} &
{\small{0.92}} &
{\small{76068}} &
{\small{75330}}\tabularnewline
\hline 
{\small{Negative Log Density}} &
{\small{222144}} &
{\small{242676}} &
{\small{1.00}} &
{\small{185748}} &
{\small{185843}}\tabularnewline
\hline 
{\small{Mean Absolute Error}} &
{\small{0.0382}} &
{\small{0.0390}} &
{\small{0.92}} &
{\small{0.0303}} &
{\small{0.0293}}\tabularnewline
\hline 
{\small{Root Mean Square Error}} &
{\small{0.0944}} &
{\small{0.0986}} &
{\small{0.77}} &
{\small{0.0701}} &
{\small{0.0677}}\tabularnewline
\hline 
\end{tabular}

@@

\begin{tabular}{|c|c|c|}
\hline 
 &
{\small{CAR model}} &
{\small{SPDE model}}\tabularnewline
\hline 
{\small{Coverage}} &
{\small{0.977}} &
{\small{0.978}}\tabularnewline
\hline 
{\small{Mean Interval Length}} &
{\small{0.135}} &
{\small{0.142}}\tabularnewline
\hline 
{\small{Median Interval Length}} &
{\small{0.040}} &
{\small{0.031}}\tabularnewline
\hline 
\end{tabular}

@@

%
\begin{table}
\caption{Predictive ability based on several predictive score criteria for
the CAR and SPDE spatial models when holding out 80\% of entire cells
of data in Minnesota. Smaller values are better, except for coverage,
where values near 0.95 are optimal.}

\begin{tabular}{|c|c|c|>{\centering}p{3cm}|c|c|}
\hline 
 &
\multicolumn{3}{c|}{{\small{Posterior mean of score}}} &
\multicolumn{2}{c|}{{\small{Score of posterior mean predictions}}}\tabularnewline
\hline 
\hline 
 &
{\small{CAR model}} &
{\small{SPDE model}} &
{\small{Posterior Prob. CAR < SPDE}} &
{\small{CAR model}} &
{\small{SPDE model}}\tabularnewline
\hline 
{\small{Brier}} &
{\small{65870}} &
{\small{65004}} &
{\small{0.08}} &
{\small{60169}} &
{\small{60020}}\tabularnewline
\hline 
{\small{Negative Log Density}} &
{\small{167947}} &
{\small{167536}} &
{\small{0.43}} &
{\small{145404}} &
{\small{145750}}\tabularnewline
\hline 
{\small{Mean Absolute Error}} &
{\small{0.0322}} &
{\small{0.0307}} &
{\small{0.08}} &
{\small{0.0246}} &
{\small{0.0240}}\tabularnewline
\hline 
{\small{Root Mean Square Error}} &
{\small{0.0800}} &
{\small{0.0771}} &
{\small{0.01}} &
{\small{0.0578}} &
{\small{0.0571}}\tabularnewline
\hline 
\end{tabular}

@@

\begin{tabular}{|c|c|c|}
\hline 
 &
{\small{CAR model}} &
{\small{SPDE model}}\tabularnewline
\hline 
{\small{Coverage}} &
{\small{0.982}} &
{\small{0.974}}\tabularnewline
\hline 
{\small{Mean Interval Length}} &
{\small{0.117}} &
{\small{0.107}}\tabularnewline
\hline 
{\small{Median Interval Length}} &
{\small{0.030}} &
{\small{0.021}}\tabularnewline
\hline 
\end{tabular}

@@

Table \ref{tab:score_tree} shows the predictive ability for Experiment
2. Here we have limited evidence (posterior probability of 0.84) that
the SPDE model is better based on the Brier score.

%
\begin{table}
\caption{Predictive ability based on several predictive score criteria for
the CAR and SPDE spatial models when holding out 5\% of trees. Smaller
values are better.}

\begin{tabular}{|c|c|c|>{\centering}p{3cm}|c|c|}
\hline 
 &
\multicolumn{3}{c|}{{\small{Posterior mean of score}}} &
\multicolumn{2}{c|}{{\small{Score of posterior mean predictions}}}\tabularnewline
\hline 
\hline 
 &
{\small{CAR model}} &
{\small{SPDE model}} &
{\small{Posterior Prob. CAR < SPDE}} &
{\small{CAR model}} &
{\small{SPDE model}}\tabularnewline
\hline 
{\small{Brier}} &
{\small{13158}} &
{\small{13143}} &
{\small{0.16}} &
{\small{13040}} &
{\small{13039}}\tabularnewline
\hline 
{\small{Negative Log Density}} &
{\small{31984}} &
{\small{31870}} &
{\small{0.02}} &
{\small{31212}} &
{\small{31229}}\tabularnewline
\hline 
\end{tabular}

@@

For the final data product, we ran the model using the CAR specification
for X iterations, discarding the first Z iterations for burn-in and
retaining Y to limit storage needs and limit the size of the data
product. Mixing was generally reasonable, but for some of the hyperparameters
was relatively slow, particularly for less common taxa. Despite this,
mixing for the variables of substantive interest, the proportions
was good. {[}Report ESS values{]}

In Figure \ref{fig:select_maps}, we show maps of estimated composition
for the full domain for several taxa of substantive interest to illustrate
the results: beech, cherry, chestnut, elm, hemlock, oak, and pine.
These maps contrast the raw data proportions, the posterior means
and posterior standard deviations as pointwise estimates of uncertainty.
Fig. \ref{fig:all_predictions} shows posterior means for all 22 (???)
taxa.

@@

\hspace{3mm}\includegraphics[scale=0.17]{2_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendRaw.pdf}\includegraphics[scale=0.17]{3_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendProp.pdf}\includegraphics[scale=0.17]{4_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendSD.pdf}

\caption{{[}THIS IS AN OLD VERSION FOR ILLUSTRATION ONLY - TO BE REPLACED WHEN
VERSION 0.3 FITS ARE DONE{]} Empirical proportions from raw data (column
1), posterior mean predictions (column 2) and posterior standard deviations
-- representing standard errors of prediction (column 3) for select
taxa.}
\end{figure}

@@

\includegraphics[scale=0.35]{5_var_tmp_paciorek_paleon_comp_0_2_output_PLScomposition_full_0_1_fits.pdf}

\caption{{[}THIS IS AN OLD VERSION FOR ILLUSTRATION ONLY - TO BE REPLACED AND
PUT IN 8 ROW X 3 COLUMN FORM WHEN VERSION 0.3 FITS ARE DONE{]} Posterior
mean predictions for all taxa over the entire domain.}

@@

While our data product is given for the full rectangular domain, including
hypothetical vegetation over water, uncertainty is large outside of
the core states with data and any analysis and inference should only
be done within those core states.

The data product is publically available at XX under YY license as
version 1.0 as of January 2015. The product is in the form of a netCDF-4
file, with dimensions x, y, and MCMC iteration. There is one variable
per taxon. The PalEON project will continue to maintain this product,
releasing new versions as additional data in Illinois, Indiana and
Ohio are digitized.

@@

Other related data products that are under development include:
\begin{itemize}
\item The gridded raw count data for the western subdomain and data aggregated
to township for the eastern subdomain. The raw data product is available
at XXX and that product provides the input data for the product described
in this work. Note that we cannot provide the raw data at point locations
because of limitations specified in the data use agreements under
which we have access to the data.
\item Gridded raw biomass estimates for Minnesota, Wisconsin, and Michigan
{[}right Simon, or also for IL and IN{]} based on the PLS data.
\item Statistically estimated biomass for Minnesota, Wisconsin, and Michigan
{[}also IL and IN?{]} using a statistical model applied to the raw
biomass estimates.
\end{itemize}
An additional drawback of the product is its focus on composition,
which does not directly tell us about vegetation structure, in particular
does not distinguish between closed forest, savanna, and prairie,
of particular note in Minnesota, Wisconsin, Illinois, and into Indiana.
The statistically-estimated biomass product mentioned just above will
directly inform questions about vegetation structure. Extensions of
that product will also estimate basal area and stem density.

>>>>

full edit

properly cite Goring stuff

@@

check journal requirements for archiving and presentation; talk to
Jody/Ann about archiving/DOI

\begin{abstract}
We present a data product of the estimated composition of tree taxa
at the time of European settlement of the United States and the statistical
methodology used to produce the product. Composition is defined as
the proportion of stems larger than approximately X cm diameter at
breast height {[}Charlie? Simon? is X roughly 10?{]} in 22 taxonomic
groupings, generally at the genus level. The data come from settlement
survey records that provide raw data that are transcribed and then
aggregated spatially, giving count data. The domain is divided into
two regions, western (Indiana through Minnesota) and eastern (Ohio
to Maine). Public Land Survey point data in the western region are
aggregated to a regular 8 km grid, while data in the eastern region,
from Town Proprietor Surveys is aggregated at the township level.
The product is based on a Bayesian statistical model fit to the count
data that estimates composition on a regular 8 km grid. The statistical
model allows us to estimate composition at locations with no data
and to smooth over noise caused by limited counts in locations with
data. Critically, it also allows us to quantify uncertainty in our
composition estimates. We expect this data product to be useful for
understanding the state of vegetation in the northeastern United States
prior to large-scale European settlement. In addition to specific
regional questions, the data product can also serve as a baseline
against which to investigate how forests and ecosystems change after
intensive settlement.
\end{abstract}

@@

Euro-American settlement and subsequent land use change occured in
a time transient fashion across North America, and surveys of vegetation
followed a similar pattern. Early surveys (from 1620 until 1825) in
the northeastern United States were areal surveys at a township level
\citep{cogbill2002forests,thompson2013four}, which we call the Town
Proprietor Survey (TPS). These surveys reporting percent composition
of major tree species using sometimes inconsistent common names. Later
surveys, after the establishment of the U.S. Public Land Survey System
(PLS) - from 1832 to 1907 - were point estimates along a regular grid,
with 1 mile spacing \citep{schulte2001original,bourdo1956review,goring2015composition}.
Survey instructions during the PLS varied through time, and by point
type. This required the application of spatially varying correction
factors \citep{cogbill2015corrections,goring2015composition} to accurately
assess stem density, basal area and biomass from the early settlement
records.

Logging, agriculture and abandonment has left an indelible mark on
forests in the northeastern United States \citep{foster1998land,goring2015composition,rhemtulla2009legacies,thompson2013four},
however most studies have studied these effects at the state level
or below \citep{rhemtulla2009historical,friedman2005regional}, and
with varying spatial resolution, from townships (36 square miles)
to forest zones of hundreds or thousands of square miles. \citet{goring2015composition}
provide a dataset across the upper Midwest that is resolved to an
8 x 8 km grid size, providing broad spatial coverage at a meso-spatial
scale that can be compared to the Forest Inventory and Analysis products
\citet{gray2012forest}. The limitation of this dataset is that it
provides only estimates of variance within cell for PLS data in the
western region, and no township estimates of error for regions surveyed
as part of the TPS.

@@

The raw data are obtained from survey records collated from across
the northeastern U.S. by a number of researchers. For the states of
Minnesota, Wisconsin, Illinois, Indiana, and Michigan (the western
subdomain), data are available at survey point locations and have
been aggregated to a regular 8 km grid in the Albers projection. (Note
that for Indiana and Illinois, at the moment trees are associated
with township centroids and then assigned to 8 km grid cells based
on the centroid but in the near future we will have point locations
available for each tree.) For the states of Ohio, Pennsylvania, New
Jersey, New York and the six New England states (the eastern subdomain),
data are aggregated at the township level. There is also data from
a single township in Quebec and a single township in northern Delaware.
Data are essentially complete in Minnesota, Wisconsin and Michigan
but data in Illinois and Indiana represent a sample of the full set
of grid cells, with survey record transcription ongoing. Data for
the remaining states are a subset of the full set of townships. Fig.
\ref{fig:domain} {[}Fig. 1 - Lyx is having a numbering problem{]}
shows the domain, indicating the grid cells and townships with data.

@@

\caption{Spatial domain, with locations with data shown in gray. Locations
are grid cells in western portion and townships in eastern portion.
Grid cells completely covered in water are white (e.g., a few locations
in Minnesota and Wisconsin).}
\end{figure}

@@

Extensive details on the data are available in \cite{Gori:etal:2015}
and the raw data are available at XXX. The aggregation into taxonomic
groups is primarily at the genus level, but in some cases of monospecific
genera, is at the species level. We model the following 22 taxa plus
an ``other hardwood'' category: Atlantic white cedar (Chamaecyparis
thyoides), Ash (Fraxinus spp.), Basswood (Tilia americana), Beech
(Fagus grandifolia), Birch (Betula spp.), Black gum/sweet gum (Nyssa
sylvatica and Liquidambar styraciflua), Cedar/juniper (what is in
here other than Juniperus and is the Juniperus only Juniperus virginiana?),
Cherry (Prunus spp.), Chestnut (Castanea dentata), Dogwood (Cornus
spp.), Elm (Ulmus spp. or Ulmus americana???) , Fir (Abies spp.),
Hemlock (Tsuga canadensis), Hickory (Carya spp.), Ironwood (Carpinus
caroliniana and Ostrya virginiana (anything else?)), Maple (Acer spp.),
Oak (Quercus spp.), Pine (Pinus spp.), Poplar/tulip poplar (Populus
spp. and Liriodendron tulipifera), Spruce (Picea spp.), Tamarack (Larix
laricina), Walnut (Juglans nigra). {[}Simon/Charlie to check my Latin
and whether I've left out anything that falls into a given common
name{]} Note that in several cases (e.g., black gum/sweet gum, ironwood,
poplar/tulip poplar, cedar/juniper), because of ambiguity in the common
tree names used by surveyors, a group represents trees from different
families and even orders. For the western subdomain we do not fit
statistical models for Atlantic white cedar and chestnut as these
have 0 and 6 trees present, respectively. The taxa grouped into other
hardwood are those for which fewer than roughly 2000 trees were present
in the dataset; however we include Atlantic white cedar explicitly
despite it only having 336 trees in the dataset.

@@

There are approximately 860,000 trees from the western subdomain and
420,000 trees from the eastern subdomain. In the western subdomain,
oak is the most common taxon and pine the second most common, while
in the eastern subdomain oak is the most common and beech the second
most common.

Our domain is a rectangle covering all of the states, in the Albers
projection (NAD 1983 Albers Great Lakes and St. Lawrence), with the
rectangle split into 8 km cells, arranged in a 296 by 180 grid of
cells, with the centroid of the cell in the southeast corner located
at (-71000, 58000). For the modeling of the western subdomain we use
the western-most 146 by 180 grid of cells. For the modeling of the
eastern subdomain we use the eastern-most 180 by 180 grid of cells
and then omit 23 cells in the north and 17 cells in the south.

@@

We compared the CAR and SPDE models using cross-validation by holding
out data from the fitting process and assessing the fit of the model
on the held-out data. This cross-validation used a subregion containing
most of Minnesota and a small amount of western Wisconsin, defined
to be the cells whose x-coordinate was less than 300,000 (this defines
a north-south line that approximately goes through Duluth, Minnesota)
and hereafter referred to as the ``Minnesota subregion'' . We used
two types of test data:
\begin{enumerate}
\item We held out all the data from 95\% of the cells in the Minnesota subregion,
with cells selected at random. This was meant to assess the ability
of the model to interpolate from a sparse set of cells/townships and
mimics the limited data in Illinois and Indiana.
\item We held out 5\% of the trees from all of the trees in the dataset
(leaving aside the held-out Minnesota subregion cells). This was meant
to assess the ability of the model to estimate the composition in
cells in which data were available. 
\end{enumerate}
Finally, in a separate sensitivity analysis we instead left out 80\%
of the cells in Minnesota subregion at random. This variation on the
first test above was meant to indicate whether our model comparison
conclusions would be robust as the digitization process for Illinois
and Indiana progresses and provides us with increasingly dense data.

@@

We ran each model for 150,000 iterations, retaining 250 after a burnin
(25,000 iterations) and subsampling to reduce storage needs.

@@

Table \ref{tab:score_cell_fivepercent} shows predictive ability for
Experiment 1: those cells in the Minnesota subregion held out of the
fitting process, for both the posterior mean predictions and the full
posterior sample. Table \ref{tab:coverage_cell_fivepercent} shows
performance of the uncertainty estimates. For the posterior distribution
over the predictive score values, the CAR model outperforms the SPDE
model, while for the posterior mean predictions, the SPDE model appears
to outperform the CAR model, but we do not have any uncertainty estimates
for this comparison. Coverage and interval lengths are similar between
the two models. From a practical perspective, based on the difference
in mean absolute error, which is on a readily interpretable scale,
the differences between the models are small.

\begin{table}
\caption{Predictive ability based on several predictive score criteria for
the CAR and SPDE spatial models when holding out 95\% of entire cells
of data in Minnesota. Smaller values are better.}

\begin{tabular}{|c|c|c|>{\centering}p{3cm}|>{\centering}p{2.5cm}|>{\centering}p{2.5cm}|}
\hline 
 &
\multicolumn{3}{c|}{{\small{Posterior mean of score}}} &
\multicolumn{2}{c|}{{\small{Score of posterior mean predictions}}}\tabularnewline
\hline 
\hline 
 &
{\small{CAR model}} &
{\small{SPDE model}} &
{\small{Posterior Prob. CAR < SPDE}} &
{\small{CAR model}} &
{\small{SPDE model}}\tabularnewline
\hline 
{\small{Brier}} &
{\small{180688}} &
{\small{186045}} &
{\small{0.98}} &
{\small{162813}} &
{\small{161559}}\tabularnewline
\hline 
{\small{Negative Log Density}} &
{\small{468397}} &
{\small{513144}} &
{\small{1.00}} &
{\small{395685}} &
{\small{395998}}\tabularnewline
\hline 
{\small{Mean Absolute Error}} &
{\small{0.0364}} &
{\small{0.0383}} &
{\small{0.98}} &
{\small{0.0275}} &
{\small{0.0269}}\tabularnewline
\hline 
{\small{Root Mean Square Error}} &
{\small{0.0897}} &
{\small{0.0959}} &
{\small{0.96}} &
{\small{0.0647}} &
{\small{0.0626}}\tabularnewline
\hline 
\end{tabular}

@@

\begin{tabular}{|c|c|c|}
\hline 
 &
{\small{CAR model}} &
{\small{SPDE model}}\tabularnewline
\hline 
{\small{Coverage}} &
{\small{0.976}} &
{\small{0.978}}\tabularnewline
\hline 
{\small{Mean Interval Length}} &
{\small{0.129}} &
{\small{0.142}}\tabularnewline
\hline 
{\small{Median Interval Length}} &
{\small{0.037}} &
{\small{0.033}}\tabularnewline
\hline 
\end{tabular}

@@

\begin{table}
\caption{Predictive ability based on several predictive score criteria for
the CAR and SPDE spatial models when holding out 80\% of entire cells
of data in Minnesota. Smaller values are better, except for coverage,
where values near 0.95 are optimal.}

\begin{tabular}{|c|c|c|>{\centering}p{3cm}|c|c|}
\hline 
 &
\multicolumn{3}{c|}{{\small{Posterior mean of score}}} &
\multicolumn{2}{c|}{{\small{Score of posterior mean predictions}}}\tabularnewline
\hline 
\hline 
 &
{\small{CAR model}} &
{\small{SPDE model}} &
{\small{Posterior Prob. CAR < SPDE}} &
{\small{CAR model}} &
{\small{SPDE model}}\tabularnewline
\hline 
{\small{Brier}} &
{\small{141546}} &
{\small{140211}} &
{\small{0.10}} &
{\small{130006}} &
{\small{129910}}\tabularnewline
\hline 
{\small{Negative Log Density}} &
{\small{356985}} &
{\small{355640}} &
{\small{0.31}} &
{\small{312509}} &
{\small{312950}}\tabularnewline
\hline 
{\small{Mean Absolute Error}} &
{\small{0.0309}} &
{\small{0.0297}} &
{\small{0.10}} &
{\small{0.0225}} &
{\small{0.0222}}\tabularnewline
\hline 
{\small{Root Mean Square Error}} &
{\small{0.0763}} &
{\small{0.0740}} &
{\small{0.04}} &
{\small{0.0531}} &
{\small{0.0528}}\tabularnewline
\hline 
\end{tabular}

@@

\begin{tabular}{|c|c|c|}
\hline 
 &
{\small{CAR model}} &
{\small{SPDE model}}\tabularnewline
\hline 
{\small{Coverage}} &
{\small{0.981}} &
{\small{0.971}}\tabularnewline
\hline 
{\small{Mean Interval Length}} &
{\small{0.112}} &
{\small{0.103}}\tabularnewline
\hline 
{\small{Median Interval Length}} &
{\small{0.028}} &
{\small{0.021}}\tabularnewline
\hline 
\end{tabular}

@@

Table \ref{tab:score_tree} shows the predictive ability for Experiment
2. Here we have evidence (posterior probability of 0.94) that the
SPDE model is better based on the Brier score.

\begin{table}
\caption{Predictive ability based on several predictive score criteria for
the CAR and SPDE spatial models when holding out 5\% of trees. Smaller
values are better.}

\begin{tabular}{|c|c|c|>{\centering}p{3cm}|c|c|}
\hline 
 &
\multicolumn{3}{c|}{{\small{Posterior mean of score}}} &
\multicolumn{2}{c|}{{\small{Score of posterior mean predictions}}}\tabularnewline
\hline 
\hline 
 &
{\small{CAR model}} &
{\small{SPDE model}} &
{\small{Posterior Prob. CAR < SPDE}} &
{\small{CAR model}} &
{\small{SPDE model}}\tabularnewline
\hline 
{\small{Brier}} &
{\small{21296}} &
{\small{21277}} &
{\small{0.06}} &
{\small{21146}} &
{\small{21143}}\tabularnewline
\hline 
{\small{Negative Log Density}} &
{\small{52199}} &
{\small{52052}} &
{\small{0.01}} &
{\small{51141}} &
{\small{51156}}\tabularnewline
\hline 
\end{tabular}

@@

For the final data product, we ran the model using the CAR specification
for 150,000 iterations, discarding the first 25,000 iterations for
burn-in and retaining 250 to limit storage needs and limit the size
of the data product. Mixing was generally reasonable, but for some
of the hyperparameters was relatively slow, particularly for less
common taxa. Despite this, mixing for the variables of substantive
interest, the proportions was good. {[}Report ESS values{]}

In Figure \ref{fig:select_maps}, we show maps of estimated composition
for the full domain for several taxa of substantive interest to illustrate
the results: beech, cherry, chestnut, elm, hemlock, oak, and pine.
These maps contrast the raw data proportions, the posterior means
and posterior standard deviations as pointwise estimates of uncertainty.
Fig. \ref{fig:all_predictions} shows posterior means for all 23 taxa.

@@

\hspace{4mm}\includegraphics[scale=0.16]{2_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendRaw.pdf}\hspace{4mm}\includegraphics[scale=0.16]{3_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendProp.pdf}\hspace{3.5mm}\includegraphics[scale=0.16]{4_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendSD.pdf}

\caption{Empirical proportions from raw data (column 1), posterior mean predictions
(column 2) and posterior standard deviations -- representing standard
errors of prediction (column 3) for select taxa.}
\end{figure}

@@

\includegraphics[scale=0.9]{5_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig3.pdf}

\caption{Posterior mean predictions for all taxa over the entire domain.}

@@

The data product is publicly available at XX under YY license as version
0.3 as of April 2015. The product is in the form of a netCDF-4 file,
with dimensions x, y, and MCMC iteration. There is one variable per
taxon. The PalEON project will continue to maintain this product,
releasing new versions as additional data in Illinois, Indiana and
Ohio are digitized.

@@

Other related data products that are under development include:
\begin{itemize}
\item The gridded raw count data for the western subdomain and data aggregated
to township for the eastern subdomain. The raw data product is available
at XXX and that product provides the input data for the product described
in this work. Note that we cannot provide the raw data at point locations
because of limitations specified in the data use agreements under
which we have access to the data.
\item Gridded raw biomass estimates for Minnesota, Wisconsin, and Michigan
based on the PLS data, with extension to Illinois and Indiana planned.
\item Statistically estimated biomass for Minnesota, Wisconsin, and Michigan
using a statistical model applied to the raw biomass estimates, with
extension to Illinois and Indiana planned.
\end{itemize}
An additional drawback of the product is its focus on composition,
which does not directly tell us about vegetation structure, in particular
does not distinguish between closed forest, savanna, and prairie,
of particular note in Minnesota, Wisconsin, Illinois, and into Indiana.
The statistically-estimated biomass product mentioned just above will
directly inform questions about vegetation structure. Extensions of
that product will also estimate basal area and stem density.
